---
title: "s1-Chemo processing"
author: "Zinab Doha"
date: "01/01/2026"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52)
```

## UMAP + Composition + QC

## Analysis design

(Figures + QC/Supplement):

- UMAP by celltype_major

- UMAP by chemo_group

- UMAP group.by="celltype_major" and split.by="chemo_group"

- Stacked barplot of celltype_major composition by chemo group

- QC tables: cells per celltype, counts per group×celltype (optional CSV)



## Load libraries
```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
```

## Set seed

```{r}
set.seed(4)
```



# Load data

```{r}
wu_chemo <- readRDS("wu_chemo.rds")
```


# UMAP: celltype_major

```{r}

p1 <- DimPlot(
  wu_chemo,
  reduction = "umap",
  group.by  = "celltype_major",
  label     = TRUE,
  repel     = TRUE,
  pt.size   = 0.25
) + ggtitle("UMAP by Celltype Major")
print(p1)
```

# UMAP: chemo_group
```{r}
p2 <- DimPlot(
  wu_chemo,
  reduction = "umap",
  group.by  = "chemo_group",
  pt.size   = 0.25
) + ggtitle("UMAP by Chemotherapy Group")
print(p2)
```
# UMAP: celltype_major split by chemo_group

```{r}
p3 <- DimPlot(
  wu_chemo,
  reduction = "umap",
  group.by  = "celltype_major",
  split.by  = "chemo_group",
  pt.size   = 0.25
) + ggtitle("UMAP: Celltype split by Chemo Group")
print(p3)
```


# Composition: stacked barplot

```{r}
comp_df <- wu_chemo@meta.data %>%
  group_by(chemo_group, celltype_major) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(chemo_group) %>%
  mutate(freq = n / sum(n))

p_stack <- ggplot(comp_df, aes(x = chemo_group, y = freq, fill = celltype_major)) +
  geom_col() +
  ylab("Proportion of Cells") +
  ggtitle("Celltype Composition by Chemotherapy Group") +
  theme_bw()
print(p_stack)
```

# QC tables (Supplement)
```{r}
qc_celltype <- as.data.frame(table(wu_chemo$celltype_major))
qc_group_celltype <- comp_df %>% select(chemo_group, celltype_major, n)

write.csv(qc_celltype, "QC_cells_per_celltype.csv", row.names = FALSE)
write.csv(qc_group_celltype, "QC_counts_per_group_celltype.csv", row.names = FALSE)

```


# sessionInfo()

```{r}
sessionInfo()
```

