---
title: "s2-Chemo_DGE"
author: "Zinab Doha"
date: "01/01/20265"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52)
```

## Differential Expression – Cancer Epithelial cells

## Analysis design

2 (Differential Expression – Cancer Epithelial cells):

FindMarkers(Chemo_TNBC vs Chemo_non_TNBC) within assigned celltype_major

Volcano plot

Top/Bottom gene sets (Top=up in TNBC; Bottom=up in non-TNBC)

DotPlot (top DE genes)

GO dotplot + cnetplot + heatplot (if packages installed)

Saved CSV: StudentX_<Celltype>_FindMarkers.csv



## Load libraries
```{r}


#BiocManager::install(c("clusterProfiler","org.Hs.eg.db","enrichplot","AnnotationDbi"))

library(Seurat)
library(dplyr)
library(tibble)
library(ggplot2)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(AnnotationDbi)


```

# Load data

```{r}
wu_chemo <- readRDS("wu_chemo.rds")
```

## Set seed

```{r}
set.seed(4)
```
# FindMarkers(Chemo_TNBC vs Chemo_non_TNBC)
```{r}
cells_use <- rownames(wu_chemo@meta.data)[
  wu_chemo$celltype_major == "Cancer Epithelial"
]

subset_ct <- subset(wu_chemo, cells = cells_use)
Idents(subset_ct) <- "chemo_group"

deg <- FindMarkers(
  subset_ct,
  ident.1 = "Chemo_TNBC",
  ident.2 = "Chemo_non_TNBC",
  test.use = "wilcox",
  min.pct = 0.1,
  logfc.threshold = 0
) %>% rownames_to_column("gene")

write.csv(deg, "Student2_CancerEpithelial_FindMarkers.csv", row.names = FALSE)
```

# Volcano plot
```{r}
# Columns
logfc_col <- if ("avg_log2FC" %in% colnames(deg)) "avg_log2FC" else "avg_logFC"
padj_col  <- if ("p_val_adj" %in% colnames(deg)) "p_val_adj" else "p_val"

deg <- deg %>%
  mutate(logFC = .data[[logfc_col]],
         padj = .data[[padj_col]],
         neglogP = -log10(padj + 1e-300))

# Volcano
ggplot(deg, aes(logFC, neglogP)) +
  geom_point(alpha = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  ggtitle("Cancer epithelial: TNBC vs non-TNBC")

# Gene sets
top_genes    <- deg %>% filter(logFC > 0.25, padj < 0.05) %>% pull(gene)
bottom_genes <- deg %>% filter(logFC < -0.25, padj < 0.05) %>% pull(gene)

# DotPlot (top 10 each direction)
features <- unique(c(
  head(top_genes, 10),
  head(bottom_genes, 10)
))
DotPlot(subset_ct, features = features, group.by = "chemo_group") +
  RotatedAxis() +
  ggtitle("Cancer epithelial – top markers")

# SYMBOL → ENTREZ
symbol_to_entrez <- function(genes) {
  AnnotationDbi::select(
    org.Hs.eg.db,
    keys = genes,
    keytype = "SYMBOL",
    columns = "ENTREZID"
  ) %>% distinct(SYMBOL, .keep_all = TRUE)
}

fc_vec <- deg$logFC
names(fc_vec) <- deg$gene

# GO enrichment + plots
run_GO <- function(genes, label) {
  if (length(genes) < 5) return(NULL)
  eg <- symbol_to_entrez(genes)$ENTREZID
  ego <- enrichGO(eg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID",
                  ont = "BP", pAdjustMethod = "BH", readable = TRUE)
  if (nrow(ego@result) > 0) {
    print(dotplot(ego) + ggtitle(paste(celltype, label)))
    print(cnetplot(ego, showCategory = 5))
    print(heatplot(ego, foldChange = fc_vec, showCategory = 5))
  }
}

run_GO(top_genes, "Top (Up in TNBC)")
run_GO(bottom_genes, "Bottom (Up in non-TNBC)")
```





# sessionInfo()

```{r}
sessionInfo()
```

