---
title: "s2-Chemo"
author: "Zinab Doha"
date: "01/01/20265"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52)
```

## Students 2: Differential Expression – Cancer Epithelial cells

## Analysis design

Students 2 (Differential Expression – Cancer Epithelial cells):

FindMarkers(Chemo_TNBC vs Chemo_non_TNBC) within assigned celltype_major

Volcano plot

Top/Bottom gene sets (Top=up in TNBC; Bottom=up in non-TNBC)

DotPlot (top DE genes)

GO dotplot + cnetplot + heatplot (if packages installed)

Saved CSV: StudentX_<Celltype>_FindMarkers.csv



## Load libraries
```{r}


#BiocManager::install(c("clusterProfiler","org.Hs.eg.db","enrichplot","AnnotationDbi"))

library(Seurat)
library(dplyr)
library(tibble)
library(ggplot2)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(AnnotationDbi)


```

# Load data

```{r}

wu_chemo <- readRDS("wu_chemo.rds")
```

## Set seed

```{r}
set.seed(4)
```

# subset Cancer Epithelial
```{r}
cells_use <- rownames(wu_chemo@meta.data)[
  wu_chemo$celltype_major == "Cancer Epithelial"
]

subset_ct <- subset(wu_chemo, cells = cells_use)
Idents(subset_ct) <- "chemo_group"
```

##Preprocess & cluster cancer epithelial cells
```{r}
library(Seurat)
library(dplyr)
library(ggplot2)

set.seed(4)

# Make sure RNA assay is active
#DefaultAssay(subset_ct) <- "RNA"

# Standard preprocessing
#subset_ct <- NormalizeData(subset_ct, verbose = FALSE)
subset_ct <- FindVariableFeatures(subset_ct, nfeatures = 3000, verbose = FALSE)
#subset_ct <- ScaleData(subset_ct, verbose = FALSE)

# Dimensional reduction
subset_ct <- RunPCA(subset_ct, npcs = 40, verbose = FALSE)

# Neighbors + clustering
subset_ct <- FindNeighbors(subset_ct, dims = 1:25)
subset_ct <- FindClusters(subset_ct, resolution = 0.09)

# UMAP
subset_ct <- RunUMAP(subset_ct, dims = 1:25)

# Check clusters
DimPlot(subset_ct, label = TRUE) +
  ggtitle("Cancer epithelial sub-clusters")

```


##Define epithelial functional marker programs
These are functional cancer epithelial programs, not lineage:
```{r}
lin_markers <- c(
 "MKI67", "TOP2A", "HMGB2", "UBE2C", "CENPF", "CCNB1",
 
 "FASN", "ACACA", "SCD", "FABP5", "ACLY", "ELOVL6",
 
  "VIM", "ZEB1", "ZEB2", "SNAI1", "FN1",
 
 "EPCAM", "KRT8", "KRT18", "KRT19",
 
   "KRT5", "KRT14","MYH11", "COL17A1", "ITGA6"
 
)


# Make sure identities are clusters
Idents(subset_ct) <- "seurat_clusters"

DotPlot(
   subset_ct,
   features = lin_markers,cols = 'RdYlBu')+
   theme_bw()+
   coord_flip()
```


```{r}
# Convert list to flat character vector
epi_markers_flat <- c(
  # Proliferative
  "MKI67", "TOP2A", "HMGB2", "UBE2C", "CENPF", "CCNB1",
  
  # Fatty Acid Metabolic
  "FASN", "ACACA", "SCD", "FABP5", "ACLY", "ELOVL6",
  
  # EMT
  "VIM", "ZEB1", "ZEB2", "SNAI1", "FN1",
  
  # Luminal
  "EPCAM", "KRT8", "KRT18", "KRT19", "MUC1", "XBP1"
)
```

```{r}
# ── STEP 1: check what identities exist in subset_ct ──────
print(levels(Idents(subset_ct)))
print(head(subset_ct@meta.data))  # see all available columns
```


##DotPlot to annotate epithelial clusters
This is the key step for manual annotation.
High MKI67/TOP2A → Proliferative epithelial

High FASN/ACACA → Fatty-acid metabolic epithelial

High CD274/IDO1 → Immune-suppressive epithelial

High VIM/ZEB1 → EMT-like epithelial

```{r}
# Make sure PlotIdent is set correctly first
Idents(subset_ct) <- "seurat_clusters"

# Then plot
p_epi <- DotPlot(
  subset_ct,
  features = epi_markers_flat,
  cols     = "RdYlBu",
  group.by = "seurat_clusters",
  dot.scale = 6,
  scale    = FALSE
) +
  RotatedAxis() +
  ggtitle("Cancer Epithelial — marker programs by subtype") +
  theme(
    axis.text.x     = element_text(size = 8),
    axis.text.y     = element_text(size = 9),
    plot.title      = element_text(face = "bold", size = 11),
    legend.position = "right"
  )

# Add group labels on top
epi_groups <- list(
  "Proliferative" = 6,
  "EMT"           = 5,
  "FA"     = 6,
  "Luminal"       = 4
)

p_epi

p_epi <- add_group_labels(p_epi, epi_groups, angle = 35)

ggsave("plots/DotPlot_Epithelial.pdf",
       plot = p_epi, width = 15, height = 6, dpi = 300)

```


##Assign epithelial subtype names to clusters


 
  
```{r}
epi_cluster_to_subtype <- c(
   "0" = "FA_Luminal",
  "1" = "FA_EMT",
  "2" = "Luminal_2",
  "3" = "Luminal_1")

subset_ct$PlotIdent <- unname(
  epi_cluster_to_subtype[
    as.character(Idents(subset_ct))
  ]
)

subset_ct$PlotIdent <- factor(
  subset_ct$PlotIdent,
  levels = c(
    "0" = "FA_Luminal",
  "1" = "FA_EMT",
  "2" = "Luminal_2",
  "3" = "Luminal_1"
  )
)




```

##UMAP visualizations
🔹 By epithelial subtype
```{r}
DimPlot(
  subset_ct,
  group.by = "PlotIdent",
  label = TRUE,
  repel = TRUE
) +
  ggtitle("Cancer epithelial subtypes")
```

##🔹 By chemo group (Sensitive vs Resistant)
```{r}
DimPlot(
  subset_ct,
  group.by = "chemo_group",
  cols = c(
    "Chemo_non_TNBC" = "#00BFC4",
    "Chemo_TNBC" = "#F8766D"
  )
) +
  ggtitle("Cancer epithelial cells by chemo response")
```

🔹 Split UMAP: subtype × chemo group
```{r}
DimPlot(
  subset_ct,
  group.by = "PlotIdent",
  split.by = "chemo_group",
  pt.size = 0.3
) +
  ggtitle("Epithelial subtypes split by chemo response")
```

##6️⃣ Composition of epithelial subtypes by chemo group
🔹 Table
```{r}
table(subset_ct$PlotIdent, subset_ct$chemo_group)
```

##🔹 Percentage barplot
```{r}
comp_df <- subset_ct@meta.data %>%
  filter(!is.na(PlotIdent), !is.na(chemo_group)) %>%
  group_by(chemo_group, PlotIdent) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(chemo_group) %>%
  mutate(freq = n / sum(n))

ggplot(comp_df, aes(x = chemo_group, y = freq, fill = PlotIdent)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Chemotherapy group",
    y = "Proportion of epithelial cells",
    title = "Epithelial subtype composition by chemo response"
  ) +
  theme_bw()
```

##7️⃣  Save

```{r}
saveRDS(
  subset_ct,
  file = "CancerEpithelial_Subclustered.rds"
)
```



# sessionInfo()

```{r}
sessionInfo()
```

