---
title: "s4-Chemo_CELLCHAT_INTRACTION"
author: "Zinab Doha"
date: "01/01/20265"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', out.width = "90%", fig.asp = 0.52)
```





## Load libraries


```{r}


#BiocManager::install(c("clusterProfiler","org.Hs.eg.db","enrichplot","AnnotationDbi"))

library(Seurat)
library(CellChat)
library(patchwork)
library(ggplot2)
library(dplyr)
library(future)

# # IMPORTANT: CellChat is unstable with multisession on Windows
# plan(sequential)
# 
# options(future.globals.maxSize = 20 * 1024^3)
# 
# # Clean memory
# gc()


```

# Load data

```{r}
epi   <- readRDS("data/CancerEpithelial_Subclustered.rds")
fib   <- readRDS("data/Fibroblast_Subclustered.rds")
tcell <- readRDS("data/T-cells_Subclustered.rds")
myelo <- readRDS("data/Myeloid_Subclustered.rds")

```

## Set seed

```{r}
set.seed(4)
```
# Harmonize metadata and cell type labels

We create one unified cell-type label used by CellChat.
PlotIdent, TcellPlotIdent, MyeloPlotIdent
```{r}
epi$CellChatType   <- paste0("epi_", epi$PlotIdent)
fib$CellChatType   <- paste0("fib_", fib$FibroSubtype)
tcell$CellChatType <- paste0("T_", tcell$TcellPlotIdent)
myelo$CellChatType <- paste0("myelo_", myelo$MyeloidSubtype)
```

#Merge all lineages into one Seurat object


```{r}
combined <- merge(
  epi,
  y = list(fib, tcell, myelo),
  add.cell.ids = c("epi", "fib", "T", "myelo")
)
```

```{r}
# Sanity check
table(combined$CellChatType)
table(combined$chemo_group)

```

##Split data into NR vs R
```{r}
obj_R <- subset(combined, subset = chemo_group == "Chemo_TNBC")      # R
obj_S  <- subset(combined, subset = chemo_group == "Chemo_non_TNBC")  # S

```


```{r}
library(CellChat)

cellchat_R <- createCellChat(
  object = GetAssayData(
    obj_R,
    assay = "RNA",
    layer = "data"   # <-- THIS is the key fix
  ),
  meta = obj_R@meta.data,
  group.by = "CellChatType"
)

cellchat_S <- createCellChat(
  object = GetAssayData(
    obj_S,
    assay = "RNA",
    layer = "data"
  ),
  meta = obj_S@meta.data,
  group.by = "CellChatType"
)
##CHEACK
cellchat_R
cellchat_S

```

```{r}
# CellChat group labels actually used
levels(cellchat_R@idents)
levels(cellchat_S@idents)

rownames(cellchat_R@net$weight)
colnames(cellchat_R@net$weight)
```

```{r}
cellchat_R@DB <- CellChatDB.human
cellchat_S@DB  <- CellChatDB.human

cellchat_R <- subsetData(cellchat_R)
cellchat_S  <- subsetData(cellchat_S)

cellchat_R <- identifyOverExpressedGenes(cellchat_R)
cellchat_S  <- identifyOverExpressedGenes(cellchat_S)

cellchat_R <- identifyOverExpressedInteractions(cellchat_R)
cellchat_S  <- identifyOverExpressedInteractions(cellchat_S) 

cellchat_R <- computeCommunProb(cellchat_R)
cellchat_S  <- computeCommunProb(cellchat_S)

cellchat_R <- computeCommunProbPathway(cellchat_R)
cellchat_S  <- computeCommunProbPathway(cellchat_S)

cellchat_R <- aggregateNet(cellchat_R)
cellchat_S  <- aggregateNet(cellchat_S)

```



```{r}
# Extract interaction weight matrices
mat_R <- cellchat_R@net$weight
mat_S  <- cellchat_S@net$weight

# Make sure ordering is identical
common_cells <- intersect(rownames(mat_R), rownames(mat_S))
mat_R <- mat_R[common_cells, common_cells]
mat_S  <- mat_S[common_cells, common_cells]

# Difference matrix: R − S
diff_mat <- mat_R - mat_S


```

##PANEL A
Heatmap: R − S interaction differences (ALL cells)
================================
```{r}

library(ComplexHeatmap)
library(circlize)

# Align matrices
mat_R <- cellchat_R@net$weight
mat_S  <- cellchat_S@net$weight

common_senders   <- intersect(rownames(mat_NR), rownames(mat_R))
common_receivers <- intersect(colnames(mat_NR), colnames(mat_R))

mat_NR <- mat_NR[common_senders, common_receivers]
mat_R  <- mat_R[common_senders, common_receivers]

diff_mat <- mat_NR - mat_R

# Symmetric color scaling
max_abs <- max(abs(diff_mat))

col_fun <- colorRamp2(
  c(-max_abs, 0, max_abs),
  c("#2C7BB6", "white", "#D7191C")  # blue=R, red=NR
)

ht_A <- Heatmap(
  diff_mat,
  name = "NR − R",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_title = "Sender (Ligand)",
  column_title = "Receiver (Receptor)",
  right_annotation = rowAnnotation(
    Total = anno_barplot(rowSums(abs(diff_mat)),
                         gp = gpar(fill = "grey40"))
  ),
  top_annotation = columnAnnotation(
    Total = anno_barplot(colSums(abs(diff_mat)),
                         gp = gpar(fill = "grey40"))
  )
)

draw(ht_A)
```
##Significant signaling pathways in EMT → T cell ranked by information flow (R vs S)

This reproduces panel c in the example figure.

```{r}
library(dplyr)
library(ggplot2)

# Extract EMT → T cell communication
df_NR <- subsetCommunication(cellchat_NR)
df_R  <- subsetCommunication(cellchat_R)

T_targets <- c("T_Naive","T_CD8 Proliferative","T_Treg")

df_NR_T <- df_NR %>%
  filter(source == "epi_Prolif_EMT",
         target %in% T_targets)

df_R_T <- df_R %>%
  filter(source == "epi_Prolif_EMT",
         target %in% T_targets)

# Compute pathway information flow
path_NR <- df_NR_T %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob)) %>%
  mutate(group = "NR")

path_R <- df_R_T %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob)) %>%
  mutate(group = "R")

path_compare <- bind_rows(path_NR, path_R)

# Plot
ggplot(path_compare,
       aes(x = reorder(pathway_name, info_flow),
           y = info_flow,
           fill = group)) +
  geom_bar(stat="identity", position="dodge") +
  coord_flip() +
  scale_fill_manual(values=c(
    "NR"="#F8766D",
    "R"="#00BFC4"
  )) +
  theme_bw() +
  labs(
    x="Signaling pathway",
    y="Information flow",
    title="EMT → T cell signaling pathways"
  )
```

###rigt and left C
```{r}
library(dplyr)
library(ggplot2)

# Extract EMT → T cell communication
df_NR <- subsetCommunication(cellchat_NR)
df_R  <- subsetCommunication(cellchat_R)

T_targets <- c("T_Naive", "T_CD8 Proliferative", "T_Treg")

df_NR_T <- df_NR %>%
  filter(source == "epi_Prolif_EMT",
         target %in% T_targets)

df_R_T <- df_R %>%
  filter(source == "epi_Prolif_EMT",
         target %in% T_targets)

# Compute pathway information flow
path_NR <- df_NR_T %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob), .groups = "drop") %>%
  mutate(group = "NR")

path_R <- df_R_T %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob), .groups = "drop") %>%
  mutate(group = "R")

# Make R negative so it goes to the LEFT
path_R <- path_R %>%
  mutate(info_flow = -info_flow)

# Combine
path_compare <- bind_rows(path_NR, path_R)

# Order pathways by total absolute signal
path_order <- path_compare %>%
  group_by(pathway_name) %>%
  summarise(total = sum(abs(info_flow)), .groups = "drop") %>%
  arrange(total) %>%
  pull(pathway_name)

path_compare$pathway_name <- factor(
  path_compare$pathway_name,
  levels = path_order
)

# Plot
ggplot(path_compare,
       aes(x = pathway_name,
           y = info_flow,
           fill = group)) +
  geom_col(width = 0.75) +
  coord_flip() +
  geom_vline(xintercept = NULL) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_fill_manual(values = c(
    "NR" = "#F8766D",   # salmon
    "R"  = "#00BFC4"    # teal
  )) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.title = element_blank()
  ) +
  labs(
    x = NULL,
    y = "Information flow",
    title = "EMT → T cell signaling pathways"
  )
```
##remove less than 0.5
```{r}
library(dplyr)
library(ggplot2)

# Extract EMT → T cell communication
df_NR <- subsetCommunication(cellchat_NR)
df_R  <- subsetCommunication(cellchat_R)

T_targets <- c("T_Naive", "T_CD8 Proliferative", "T_Treg")

df_NR_T <- df_NR %>%
  filter(source == "epi_Prolif_EMT",
         target %in% T_targets)

df_R_T <- df_R %>%
  filter(source == "epi_Prolif_EMT",
         target %in% T_targets)

# Compute pathway information flow
path_NR <- df_NR_T %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob), .groups = "drop") %>%
  mutate(group = "NR")

path_R <- df_R_T %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob), .groups = "drop") %>%
  mutate(group = "R")

# Keep only pathways with information flow >= 0.5 in either group
path_keep <- bind_rows(path_NR, path_R) %>%
  group_by(pathway_name) %>%
  summarise(max_flow = max(info_flow), .groups = "drop") %>%
  filter(max_flow >= 0.5) %>%
  pull(pathway_name)

path_NR <- path_NR %>% filter(pathway_name %in% path_keep)
path_R  <- path_R  %>% filter(pathway_name %in% path_keep)

# Make R negative so it goes LEFT
path_R <- path_R %>%
  mutate(info_flow = -info_flow)

# Combine
path_compare <- bind_rows(path_NR, path_R)

# Order pathways by total absolute signal
path_order <- path_compare %>%
  group_by(pathway_name) %>%
  summarise(total = sum(abs(info_flow)), .groups = "drop") %>%
  arrange(total) %>%
  pull(pathway_name)

path_compare$pathway_name <- factor(
  path_compare$pathway_name,
  levels = path_order
)

# Plot
ggplot(path_compare,
       aes(x = pathway_name,
           y = info_flow,
           fill = group)) +
  geom_col(width = 0.75) +
  coord_flip() +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_y_continuous(labels = abs) +
  scale_fill_manual(values = c(
    "NR" = "#F8766D",
    "R"  = "#00BFC4"
  )) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.title = element_blank()
  ) +
  labs(
    x = NULL,
    y = "Information flow",
    title = "EMT → T cell signaling pathways"
  )
```


##This shows which pathways dominate EMT→T signaling in R vs S.
Panel C
EMT → T cell ligand–receptor pairs (NR vs R)

You asked for the y-axis format:

T_CD8 Proliferative (NR)
T_CD8 Proliferative (R)
T_Naive (NR)
T_Naive (R)
T_Treg (NR)
T_Treg (R)
Step 1 — Prepare data
```{r}

df_NR_T$group <- "NR"
df_R_T$group  <- "R"

df_plot <- bind_rows(df_NR_T, df_R_T)

df_plot$target_group <- paste(df_plot$target, "(", df_plot$group, ")")

#Step 2 — Select strongest interactions
top_pairs <- df_plot %>%
  group_by(interaction_name) %>%
  summarise(score = max(prob)) %>%
  arrange(desc(score)) %>%
  slice_head(n=20)

df_plot <- df_plot %>%
  filter(interaction_name %in% top_pairs$interaction_name)
#Step 3 — Bubble plot (like panel d)
ggplot(df_plot,
       aes(x = target_group,
           y = interaction_name,
           size = prob,
           color = prob)) +
  geom_point() +
  scale_color_gradient(
    low = "#2C7BB6",
    high = "#D7191C"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle=45, hjust=1),
    axis.text.y = element_text(size=7)
  ) +
  labs(
    x="T cell subtype",
    y="Ligand–Receptor pair",
    title="EMT epithelial → T cell interactions"
  )
```


##EMT → MYELOID PATHWAY BARPLOT
```{r}
library(dplyr)
library(ggplot2)

# Extract EMT → Myeloid communication
df_NR <- subsetCommunication(cellchat_NR)
df_R  <- subsetCommunication(cellchat_R)

myelo_targets <- c("myelo_cDC1", "myelo_MDSCs", "myelo_TAMs")

df_NR_M <- df_NR %>%
  filter(source == "epi_Prolif_EMT",
         target %in% myelo_targets)

df_R_M <- df_R %>%
  filter(source == "epi_Prolif_EMT",
         target %in% myelo_targets)

# Compute pathway information flow
path_NR <- df_NR_M %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob), .groups = "drop") %>%
  mutate(group = "NR")

path_R <- df_R_M %>%
  group_by(pathway_name) %>%
  summarise(info_flow = sum(prob), .groups = "drop") %>%
  mutate(group = "R")

# Keep only pathways with information flow >= 0.5 in either group
path_keep <- bind_rows(path_NR, path_R) %>%
  group_by(pathway_name) %>%
  summarise(max_flow = max(info_flow), .groups = "drop") %>%
  filter(max_flow >= 0.5) %>%
  pull(pathway_name)

path_NR <- path_NR %>% filter(pathway_name %in% path_keep)
path_R  <- path_R  %>% filter(pathway_name %in% path_keep)

# Make R negative so it goes LEFT
path_R <- path_R %>%
  mutate(info_flow = -info_flow)

# Combine
path_compare <- bind_rows(path_NR, path_R)

# Order pathways by total absolute signal
path_order <- path_compare %>%
  group_by(pathway_name) %>%
  summarise(total = sum(abs(info_flow)), .groups = "drop") %>%
  arrange(total) %>%
  pull(pathway_name)

path_compare$pathway_name <- factor(
  path_compare$pathway_name,
  levels = path_order
)

# Plot
ggplot(path_compare,
       aes(x = pathway_name,
           y = info_flow,
           fill = group)) +
  geom_col(width = 0.75) +
  coord_flip() +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_y_continuous(labels = abs) +
  scale_fill_manual(values = c(
    "NR" = "#F8766D",
    "R"  = "#00BFC4"
  )) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.title = element_blank()
  ) +
  labs(
    x = NULL,
    y = "Information flow",
    title = "EMT → Myeloid signaling pathways"
  )

```

##EMT → MYELOID BUBBLE PLOT
```{r}
library(dplyr)
library(ggplot2)

# Add group labels
df_NR_M$group <- "NR"
df_R_M$group  <- "R"

# Combine
df_plot <- bind_rows(df_NR_M, df_R_M)

# Create x-axis labels
df_plot$target_group <- paste(df_plot$target, "(", df_plot$group, ")")

# Select strongest interactions (top 20)
top_pairs <- df_plot %>%
  group_by(interaction_name) %>%
  summarise(score = max(prob), .groups = "drop") %>%
  arrange(desc(score)) %>%
  slice_head(n = 20)

df_plot <- df_plot %>%
  filter(interaction_name %in% top_pairs$interaction_name)

# Optional: order x-axis manually
df_plot$target_group <- factor(
  df_plot$target_group,
  levels = c(
    "myelo_cDC1 ( NR )", "myelo_cDC1 ( R )",
    "myelo_MDSCs ( NR )", "myelo_MDSCs ( R )",
    "myelo_TAMs ( NR )", "myelo_TAMs ( R )"
  )
)

# Bubble plot
ggplot(df_plot,
       aes(x = target_group,
           y = interaction_name,
           size = prob,
           color = prob)) +
  geom_point() +
  scale_color_gradient(
    low = "#2C7BB6",
    high = "#D7191C"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 7),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(
    x = "Myeloid subtype",
    y = "Ligand–Receptor pair",
    title = "EMT epithelial → Myeloid interactions"
  )
```



##only epi a sender
```{r}
epi_senders <- grep("^epi_", levels(cellchat_NR@idents), value = TRUE)
epi_senders


mat_NR <- cellchat_NR@net$weight

# Keep epithelial senders (rows), all receivers (columns)
mat_NR_epi <- mat_NR[epi_senders, , drop = FALSE]

dim(mat_NR_epi)


```
##plot
```{r}
library(ComplexHeatmap)
library(circlize)

col_fun_NR <- colorRamp2(
  c(0, max(mat_NR_epi)),
  c("white", "#D7191C")   # stronger = red
)

ht_NR_epi <- Heatmap(
  mat_NR_epi,
  name = "NR interaction\nstrength",
  col = col_fun_NR,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  row_title = "Epithelial sender programs",
  column_title = "Receiver (TME cell types)",

  row_names_gp = gpar(fontsize = 11),
  column_names_gp = gpar(fontsize = 10),
  row_names_rot = 0,
  column_names_rot = 45,

  row_names_max_width = unit(6, "cm"),
  column_names_max_height = unit(6, "cm"),

  right_annotation = rowAnnotation(
    Total = anno_barplot(
      rowSums(mat_NR_epi),
      gp = gpar(fill = "grey40"),
      border = FALSE
    )
  ),

  top_annotation = columnAnnotation(
    Total = anno_barplot(
      colSums(mat_NR_epi),
      gp = gpar(fill = "grey40"),
      border = FALSE
    )
  ),

  heatmap_legend_param = list(
    title = "Interaction\nstrength"
  )
)


##save
pdf(
  "NR_Epithelial_to_TME_CellChat_heatmap.pdf",
  width = 10,
  height = 8,
  useDingbats = FALSE
)

draw(ht_NR_epi,
     heatmap_legend_side = "right",
     annotation_legend_side = "right")

dev.off()

```

##suplementry
```{r}
# ══════════════════════════════════════════════════════════
# SUPPLEMENTARY TABLES — FA_EMT signaling to T cell & Myeloid
# ══════════════════════════════════════════════════════════

dir.create("supplementary", showWarnings = FALSE)

# ── 1. TOP PATHWAYS: FA_EMT → T cell ──────────────────────
supp_path_T <- bind_rows(
  df_R_T %>% group_by(pathway_name) %>%
    summarise(info_flow = sum(prob), .groups = "drop") %>%
    mutate(group = "R"),
  df_S_T %>% group_by(pathway_name) %>%
    summarise(info_flow = sum(prob), .groups = "drop") %>%
    mutate(group = "S")
) %>%
  arrange(group, desc(info_flow))

write.csv(supp_path_T,
          "supplementary/SupplTable_Pathways_FAEMT_Tcell.csv",
          row.names = FALSE)

# ── 2. TOP L-R INTERACTIONS: FA_EMT → T cell ──────────────
supp_lr_T <- bind_rows(df_R_T, df_S_T) %>%
  select(source, target, interaction_name,
         pathway_name, prob, group) %>%
  arrange(group, desc(prob))

write.csv(supp_lr_T,
          "supplementary/SupplTable_LR_FAEMT_Tcell.csv",
          row.names = FALSE)

# ── 3. TOP PATHWAYS: FA_EMT → Myeloid ─────────────────────
supp_path_M <- bind_rows(
  df_R_M %>% group_by(pathway_name) %>%
    summarise(info_flow = sum(prob), .groups = "drop") %>%
    mutate(group = "R"),
  df_S_M %>% group_by(pathway_name) %>%
    summarise(info_flow = sum(prob), .groups = "drop") %>%
    mutate(group = "S")
) %>%
  arrange(group, desc(info_flow))

write.csv(supp_path_M,
          "supplementary/SupplTable_Pathways_FAEMT_Myeloid.csv",
          row.names = FALSE)

# ── 4. TOP L-R INTERACTIONS: FA_EMT → Myeloid ─────────────
supp_lr_M <- bind_rows(df_R_M, df_S_M) %>%
  select(source, target, interaction_name,
         pathway_name, prob, group) %>%
  arrange(group, desc(prob))

write.csv(supp_lr_M,
          "supplementary/SupplTable_LR_FAEMT_Myeloid.csv",
          row.names = FALSE)

message("4 supplementary CSV files saved to supplementary/")
```


# sessionInfo()

```{r}
sessionInfo()
```

