---
title: "S5_BULK_validation"
output: html_document
date: "2026-02-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



##Load cleaned datasets
```{r}
# ── STEP 1: Update fastmap and RSQLite first ──────────────
install.packages("fastmap")
install.packages("RSQLite")

# ── STEP 2: Restart R session ─────────────────────────────
# In RStudio: Session → Restart R
# OR run:
.rs.restartR()
```
```{r}
# ── STEP 1: Check where fastmap is installed ──────────────
find.package("fastmap")
.libPaths()  # shows all library paths R is searching

# ── STEP 2: Install fastmap to the correct user library ───
install.packages("fastmap", 
                 lib = .libPaths()[1],  # force install to first path
                 type = "binary")       # binary is faster on Windows

# ── STEP 3: Verify the version in the correct location ────
packageVersion("fastmap")
```
```{r}
# ── Find and remove old fastmap ───────────────────────────
# Check all installed versions
installed.packages()[,"Package"] |> 
  (\(x) x[x == "fastmap"])()

# Remove from ALL library paths
remove.packages("fastmap")

# Reinstall fresh
install.packages("fastmap", type = "binary")

# Restart R completely
.rs.restartR()
```
```{r}
# ── Verify then load ──────────────────────────────────────
packageVersion("fastmap")   # must show 1.2.0 or higher
library(GSVA)
```



```{r}
# install.packages("BiocManager")
# BiocManager::install("GSVA")




library(dplyr)
library(tibble)
library(ggplot2)
library(pheatmap)
library(GSVA)
library(msigdbr)
library(dplyr)


 gse58812 <- readRDS("gse58812_clean.rds")

expr <- gse58812$expr
meta <- gse58812$meta

dim(expr)
table(meta$response)



```



##Get Hallmark Gene Sets
```{r}
hallmark <- msigdbr(
  species = "Homo sapiens",
  category = "H"
)

hallmark_list <- hallmark %>%
  split(x = .$gene_symbol, f = .$gs_name)


```

##Select Relevant Signatures
```{r}


selected_sets <- hallmark_list[c(
   "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION",
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_FATTY_ACID_METABOLISM",
  "HALLMARK_GLYCOLYSIS",
  "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
   "HALLMARK_CHOLESTEROL_HOMEOSTASIS",
  "HALLMARK_XENOBIOTIC_METABOLISM",
  "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY",
  "REACTOME_FATTY_ACID_METABOLISM",
  "HALLMARK_MTORC1_SIGNALING",
  "HALLMARK_HYPOXIA",
  "HALLMARK_APICAL_JUNCTION",
  "HALLMARK_HEDGEHOG_SIGNALING",
  "HALLMARK_KRAS_SIGNALING_DN"
)]


#Check overlap:

lapply(selected_sets, function(x) sum(x %in% rownames(expr)))
```

##run GSVA:
##Log-transform + Gaussian

This is most common and stable.
```{r}


expr_log <- log2(expr + 1)

gsva_scores <- gsva(
  expr = as.matrix(expr_log),
  gset.idx.list = selected_sets,
  method = "gsva",
  kcdf = "Gaussian",
  verbose = FALSE
)

#head(gsva_scores)
```
##add score
```{r}
# ----------------------------
# METABOLIC HALLMARKS
# ----------------------------

meta$Glycolysis_score <- gsva_scores["HALLMARK_GLYCOLYSIS", ]

meta$CHOL_score <- gsva_scores["HALLMARK_CHOLESTEROL_HOMEOSTASIS", ]

meta$MTORC1_score <- gsva_scores["HALLMARK_MTORC1_SIGNALING", ]

meta$Hypoxia_score <- gsva_scores["HALLMARK_HYPOXIA", ]


# ----------------------------
# PLASTICITY / EMT HALLMARKS
# ----------------------------

meta$EMT_score <- gsva_scores["HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", ]

meta$ApicalJunction_score <- gsva_scores["HALLMARK_APICAL_JUNCTION", ]



meta$Hedgehog_score <- gsva_scores["HALLMARK_HEDGEHOG_SIGNALING", ]

meta$KRAS_DN_score <- gsva_scores["HALLMARK_KRAS_SIGNALING_DN", ]


```

##creat combin axies
```{r}
# Core Metabolic Axis
meta$FA_axis <- rowMeans(
  meta[, c(
     "HALLMARK_FATTY_ACID_METABOLISM",
    "Glycolysis_score",
    "CHOL_score",
    "MTORC1_score",
    "Hypoxia_score"
  )]
)

# Plasticity / EMT Axis
meta$EMT_axis <- rowMeans(
  meta[, c(
    "EMT_score",
    "ApicalJunction_score",
    "Hedgehog_score",
    "KRAS_DN_score"
  )]
)


# Core co_express Axis
meta$FA_EMT_axis <- rowMeans(
  meta[, c(
     "HALLMARK_FATTY_ACID_METABOLISM",
    "Glycolysis_score",
    "CHOL_score",
    "MTORC1_score",
    "Hypoxia_score",
     "EMT_score",
    "ApicalJunction_score",
    "Hedgehog_score",
    "KRAS_DN_score"
  )]
)

```


##test resectence
```{r}
meta$response <- factor(meta$response,
                        levels = c("Sensitive","Resistant"))

wilcox.test(FA_axis ~ response, data = meta)
wilcox.test(EMT_axis ~ response, data = meta)

wilcox.test(FA_EMT_axis ~ response, data = meta)

```

```{r}
library(tidyr)
library(ggplot2)

meta_long <- meta %>%
  select(response, FA_axis, EMT_axis, FA_EMT_axis) %>%
  pivot_longer(
    cols = -response,
    names_to = "Axis",
    values_to = "Score"
  )

ggplot(meta_long,
       aes(x = response,
           y = Score,
           fill = response)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  facet_wrap(~Axis, scales = "free_y") +
  scale_fill_manual(values = c(
    "Sensitive" = "#00BFC4",
    "Resistant" = "#F8766D"
  )) +
  labs(
    x = NULL,
    y = "GSVA Score",
    title = "Metabolic and Plasticity Axes in Resistant vs Sensitive"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )


```
##SAVE
```{r}
# ── ASSIGN TO OBJECT FIRST ────────────────────────────────
p_gsva_box <- ggplot(meta_long,
       aes(x = response,
           y = Score,
           fill = response)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  facet_wrap(~Axis, scales = "free_y") +
  scale_fill_manual(values = c(
    "Sensitive" = "#00BFC4",
    "Resistant" = "#F8766D"
  )) +
  labs(
    x     = NULL,
    y     = "GSVA Score",
    title = "Metabolic and Plasticity Axes in Resistant vs Sensitive"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text      = element_text(face = "bold"),
    plot.title      = element_text(hjust = 0.5, face = "bold")
  )

# ── SAVE HIGH RESOLUTION PNG ──────────────────────────────
ggsave(
  filename = "plots/GSVA_Boxplot_Metabolic_Plasticity.png",
  plot     = p_gsva_box,
  width    = 8,
  height   = 6,
  dpi      = 600,        # publication quality
  units    = "in",
  bg       = "white"     # white background
)

# ── ALSO SAVE AS PDF (vector — best for manuscript) ───────
ggsave(
  filename = "plots/GSVA_Boxplot_Metabolic_Plasticity.pdf",
  plot     = p_gsva_box,
  width    = 8,
  height   = 6,
  units    = "in"
)

message("Saved PNG at 600 DPI and PDF to plots/")
```


```{r}
cor.test(meta$Metabolic_axis, meta$Plasticity_axis)


glm(response ~ Metabolic_axis + Plasticity_axis,
    data = meta,
    family = binomial)

```
##survaival
```{r}
library(survival)
library(survminer)
library(dplyr)

meta_surv <- meta %>%
  filter(!is.na(surv_time_days),
         !is.na(surv_event),
         !is.na(FA_axis),
         !is.na(EMT_axis),
         !is.na(FA_EMT_axis))




```

##FA_METABOLIC AXIS

```{r}
median_met <- median(meta_surv$FA_axis)

meta_surv$FA_group <- ifelse(
  meta_surv$FA_axis >= median_met,
  "High_FA_Score", "Low_FA_Score"
)

meta_surv$FA_group <- factor(
  meta_surv$FA_group,
  levels = c("Low_FA_Score","High_FA_Score")
)




```


```{r}
# ── STEP 1: Install missing package ───────────────────────
install.packages("litedown")

# ── STEP 2: Restart R session ─────────────────────────────
.rs.restartR()
```


```{r}
fit_met <- survfit(
  Surv(surv_time_days, surv_event) ~ FA_group,
  data = meta_surv
)

ggsurvplot(
  fit_met,
  data = meta_surv,
  pval = TRUE,
  risk.table = TRUE,
  conf.int = TRUE,
  palette = c("#00BFC4","#F8766D"),
  xlab = "Time (days)",
  ylab = "Survival probability",
  title = "Survival by FA Metabolic Axis",
  legend.title = "FA Metabolic Axis"
)

```
```{r}
cox_met <- coxph(
  Surv(surv_time_days, surv_event) ~ Metabolic_axis,
  data = meta_surv
)

summary(cox_met)

```


##PLASTICITY AXIS
```{r}
median_plast <- median(meta_surv$Plasticity_axis)

meta_surv$Plasticity_group <- ifelse(
  meta_surv$Plasticity_axis >= median_plast,
  "High_Plasticity", "Low_Plasticity"
)

meta_surv$Plasticity_group <- factor(
  meta_surv$Plasticity_group,
  levels = c("Low_Plasticity","High_Plasticity")
)



```

##Kaplan–Meier
```{r}


fit_plast <- survfit(
  Surv(surv_time_days, surv_event) ~ Plasticity_group,
  data = meta_surv
)

ggsurvplot(
  fit_plast,
  data = meta_surv,
  pval = TRUE,
  risk.table = TRUE,
  conf.int = TRUE,
  palette = c("#00BFC4","#F8766D"),
  xlab = "Time (days)",
  ylab = "Survival probability",
  title = "Survival by Plasticity Axis",
  legend.title = "Plasticity Axis"
)



```

```{r}
cox_plast <- coxph(
  Surv(surv_time_days, surv_event) ~ Plasticity_axis,
  data = meta_surv
)

summary(cox_plast)

```

```{r}
cox_both <- coxph(
  Surv(surv_time_days, surv_event) ~ Metabolic_axis + Plasticity_axis,
  data = meta_surv
)

summary(cox_both)

```
##Forest Plot of Cox Model (Clean Visualization)-metabolic

```{r}
library(survminer)

ggforest(cox_both, data = meta_surv)

```

#Predict Risk Score from Cox Model/Visualize Risk Distribution

The Cox model estimates a linear predictor:This shows if high-risk patients overlap with events.

```{r}
meta_surv$Risk_score <- predict(cox_both, type = "lp")


ggplot(meta_surv, aes(x = Risk_score, fill = factor(surv_event))) +
  geom_density(alpha = 0.4) +
  theme_classic() +
  labs(fill = "Event")

```


##Stratify Into High vs Low Risk/ Kaplan–Meier by Predicted Risk

Median split (simple and standard):
```{r}
median_risk <- median(meta_surv$Risk_score)

meta_surv$Risk_group <- ifelse(
  meta_surv$Risk_score >= median_risk,
  "High_Risk", "Low_Risk"
)

meta_surv$Risk_group <- factor(meta_surv$Risk_group,
                                levels = c("Low_Risk","High_Risk"))


##Kaplan–Meier by Predicted Risk
fit_risk <- survfit(
  Surv(surv_time_days, surv_event) ~ Risk_group,
  data = meta_surv
)

ggsurvplot(
  fit_risk,
  data = meta_surv,
  pval = TRUE,
  risk.table = TRUE,
  conf.int = TRUE,
  palette = c("#00BFC4","#F8766D"),
  title = "Survival by Predicted Risk (Cox Model)"
)


```
##Time-Dependent ROC (Model Performance)
```{r}
install.packages("timeROC")
library(timeROC)

```





